---
title: "BIOS 707 Project"
author: "Costa Stavrianidis"
date: "2022-11-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(VIM)
library(missMDA)
```

## Reading in Dataset
```{r}
dat <- haven::read_xpt("LLCP2021.XPT")
```

## Data Cleaning
```{r}
# Variables are initially chosen based on domain knowledge of risk factors for heart disease
# Many variables are also redundant, thus one will be chosen for that feature
survey1 <- dat %>% select(`_MICHD`, ADDEPEV3, MENTHLTH, `_RFHLTH`,
                          PHYSHLTH,`_HCVU652`, MEDCOST1, `_RFHYPE6`,
                          `_RFCHOL3`, CVDSTRK3, DIABETE4, `_SEX`,
                          `_EDUCAG`, INCOME3, `_BMI5`, `_SMOKER3`,
                          AVEDRNK3, `_TOTINDA`, DRADVISE,
                          `_AGE_G`, `_FRTLT1A`, `_VEGLT1A`) %>% 
  rename(Gen_Health = `_RFHLTH`, Phys_Health = PHYSHLTH,
         Ment_Health = MENTHLTH, Med_Cost = MEDCOST1,
         Stroke = CVDSTRK3, Diabetes = DIABETE4,
         Sex = `_SEX`, Income = INCOME3,
         Avg_Drinks = AVEDRNK3, Sodium_Intake = DRADVISE,
         Health_Cov = `_HCVU652`, BP = `_RFHYPE6`,
         Chol = `_RFCHOL3`, Heart_Disease = `_MICHD`,
         Age_Groups = `_AGE_G`, BMI = `_BMI5`,
         Education = `_EDUCAG`, Smoking = `_SMOKER3`,
         Fruit = `_FRTLT1A`, Veggies = `_VEGLT1A`,
         Exercise = `_TOTINDA`, Depression = ADDEPEV3) %>% 
  mutate(Heart_Disease = recode(Heart_Disease, '2' = 0), 
         Depression = recode(Depression, '2' = 0, '7' = as.numeric(NA), '9' = as.numeric(NA)),
         Ment_Health = recode(Ment_Health, '88' = 0, '77' = as.numeric(NA), '99' = as.numeric(NA)),
         Gen_Health = recode(Gen_Health, '2' = 0, '3' = as.numeric(NA)),
         Phys_Health = recode(Phys_Health, '88' = 0, '77' = as.numeric(NA), '99' = as.numeric(NA)),
         Health_Cov = recode(Health_Cov, '2' = 0, '9' = as.numeric(NA)),
         Med_Cost = recode(Med_Cost, '2' = 0,
                           '7' = as.numeric(NA), '9' = as.numeric(NA)),
         BP = recode(BP, '1' = 0, '2' = 1, '9' = as.numeric(NA)),
         Chol = recode(Chol, '1' = 0, '2' = 1, '9' = as.numeric(NA)),
         Stroke = recode(Stroke, '1' = 1, '2' = 0,
                         '7' = as.numeric(NA), '9' = as.numeric(NA)),
         Diabetes = recode(Diabetes, '2' = 1, '3' = 0, '4' = 2,
                           '1' = 1, '7' = as.numeric(NA),
                           '9' = as.numeric(NA)),
         Education = recode(Education, '9' = as.numeric(NA)),
         Income = recode(Income, '77' = as.numeric(NA), '99' = as.numeric(NA)),
         BMI = BMI/100,
         Smoking = recode(Smoking, '4' = 1, '3' = 2, '2' = 3, '1' = 4, '9' = as.numeric(NA)),
         Avg_Drinks = recode(Avg_Drinks,'88' = 0, '77' = as.numeric(NA), '99' = as.numeric(NA)),
         Exercise = recode(Exercise, '2' = 0, '9' = as.numeric(NA)),
         Sodium_Intake = recode(Sodium_Intake, '2' = 0, '7' = as.numeric(NA), '9' = as.numeric(NA)),
         Fruit = recode(Fruit, '2' = 0, '9' = as.numeric(NA)),
         Veggies = recode(Veggies, '2' = 0, '9' = as.numeric(NA)))

glimpse(survey1)
```

## Variable Descriptions

Heart_Disease: Respondents that have ever reported having coronary heart disease (CHD) or myocardial infarction (MI) \
0 - did not report \
1 - reported \
NA - not asked or missing

Depression: (Ever told) you that you have a depressive disorder, including depression, major depression, dysthymia, or minor depression? \
0 - no \
1 - yes \
NA - don't know, refused \

Ment_Health: Now thinking about your mental health, which includes stress, depression, and problems with emotions, for how many days during the past 30 days was your mental health not good? \
0:30 - # of days \
NA - don't know, refused

Gen_Health: Adults with good or better general health \
0 - fair or poor health \
1 - good or better health \
NA - don't know, not sure or refused/missing

Phys_Health: Now thinking about your physical health, which includes physical illness and injury, for how many days during the past 30 days was your physical health not good? \
0:30 - # days \
NA - don't know, refused, missing

Health_Cov: Respondents aged 18-64 who have any form of health care coverage \
0 - do not have \
1 - have \
NA - don't know, refused, missing

Med_Cost: Was there a time in the past 12 months when you needed to see a doctor but could not because of cost? \
0 - no \
1 - yes \
NA - don't know, refused, missing

BP: Adults who have been told they have high blood pressure by a doctor, nurse, or other health professional \
0 - no \
1 - yes \
NA - don't know, refused, missing

Chol: Adults who have had their cholesterol checked and have been told by a doctor, nurse, or other health professional that it was high \
0 - no \
1 - yes \
NA - don't know, refused, missing

Stroke: (Ever told) you had a stroke \
0 - no \
1 - yes \
NA - don't know, refused

Diabetes: (Ever told) you have diabetes (If "Yes" and respondent is female, ask "Was this only when you were pregnant?". If Respondent says pre-diabetes or borderline diabetes, use response code 4.) \
0 - no \
1 - yes \
2 - pre-diabetes or borderline \
NA - don't know, refused, missing

Sex: Indicate sex of respondent \
1 - male \
2 - female

Education: Level of education completed \
1 - did not graduate highschool \
2 - graduated highschool \
3 - attended college or technical school \
4 - graduated from college or technical school \
NA - don't know, refused, missing

Income: Is your annual household income from all sources: (If respondent refuses at any income level, code "Refused.") \
1 - <$10,000 \
2 - $10,000:$14,999 \
3 - $15,000:$19,999 \
4 - $20,000:$24,999 \
5 - $25,000:$34,999 \
6 - $35,000:$49,999 \
7 - $50,000:$74,999 \
8 - $75,000:$99,999 \
9 - $100,000:$149,999 \
10 - $150,000:$199,000 \
11 - $200,000+ \
NA - don't know, refused, missing

BMI: Body Mass Index (BMI) \
1:99.99 - bmi\
NA - don't know, refused, missing \

Smoking: Four-level smoker status: Everyday smoker, Someday smoker, Former smoker, Non-smoker \
1 - never smoked \
2 - former smoker \
3 - now smokes some days \
4 - now smokes every day \
NA - don't know, refused, missing

Avg_Drinks: One drink is equivalent to a 12-ounce beer, a 5-ounce glass of wine, or a drink with one shot of liquor. During the past 30 days, on the days when you drank, about how many drinks did you drink on the average? (A 40 ounce beer would count as 3 drinks, or a cocktail drink with 2 shots would count as 2 drinks.) \
1:76 - # of drinks \
NA - don't know, refused, missing

Exercise: Adults who reported doing physical activity or exercise during the past 30 days other than their regular job \
0 - did not have physical activity or exercise \
1 - had physical activity or exercise \
NA - don't know, refused, missing

Sodium_Intake: Has a doctor or other health professional ever advised you to reduce sodium or salt intake? \
0 - no \
1 - yes \
NA - don't know, refused, missing

Age_65: Age categories \
1 - age 18:24 \
2 - age 25:34 \
3 - age 35:44 \
4 - age 45:54 \
5 - age 55:64 \
6 - age 65+ \
NA - don't know, refused, missing

Fruit: Consume Fruit 1 or more times per day \
0 - less than one time per day \
1 - one or more times per day \
NA - don't know, refused, missing

Veggies: Consume Vegetables 1 or more times per day \
0 - less than one time per day \
1 - one or more times per day \
NA - don't know, refused, missing

## Exploratory Data Analysis
```{r, warning=FALSE}
# Missingness
colSums(is.na(survey1)) / nrow(survey1)

# Drop variables with >80% missingness
survey2 <- survey1 %>% select(-Sodium_Intake)

# Impute continuous BMI variable with mean imputation
mean_bmi <- mean(survey2$BMI, na.rm = T)
survey2 <- survey2 %>% mutate(BMI = ifelse(is.na(BMI), mean_bmi, BMI))

# Impute categorical variables with sequential, random hot-deck imputation
survey2 <- hotdeck(survey2)

# Checking for duplicates in original dataset
dat$SEQNO <- as.factor(dat$SEQNO)
duplicates <- dat %>% group_by(SEQNO, `_STATE`) %>%
  summarize(count = n()) %>% filter(count > 1)
dat2 <- dat %>% select(-c(`_STATE`:`_PSU`))
table(sapply(dat2, class))
# No duplicates found
```
