---
title: "BIOS 707 Project"
author: "Costa Stavrianidis"
date: "2022-11-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(VIM)
library(caTools)
library(visreg)
library(caret)
library(MASS)
library(boot)
library(ROCR)
library(gmodels)
```

## Reading in Dataset
```{r}
dat <- haven::read_xpt("LLCP2021.XPT")
```

## Data Cleaning
```{r}
# Variables are initially chosen based on domain knowledge of risk factors for heart disease
# Many variables are also redundant, thus one will be chosen for that feature
survey1 <- dat %>% dplyr::select(`_MICHD`, ADDEPEV3, MENTHLTH, `_RFHLTH`,
                          PHYSHLTH,`_HCVU652`, MEDCOST1, `_RFHYPE6`,
                          `_RFCHOL3`, CVDSTRK3, DIABETE4, `_SEX`,
                          `_EDUCAG`, INCOME3, `_BMI5`, `_SMOKER3`,
                          AVEDRNK3, `_TOTINDA`, DRADVISE,
                          `_AGE_G`, `_FRTLT1A`, `_VEGLT1A`) %>% 
  rename(Gen_Health = `_RFHLTH`, Phys_Health = PHYSHLTH,
         Ment_Health = MENTHLTH, Med_Cost = MEDCOST1,
         Stroke = CVDSTRK3, Diabetes = DIABETE4,
         Sex = `_SEX`, Income = INCOME3,
         Avg_Drinks = AVEDRNK3, Sodium_Intake = DRADVISE,
         Health_Cov = `_HCVU652`, BP = `_RFHYPE6`,
         Chol = `_RFCHOL3`, Heart_Disease = `_MICHD`,
         Age_Groups = `_AGE_G`, BMI = `_BMI5`,
         Education = `_EDUCAG`, Smoking = `_SMOKER3`,
         Fruit = `_FRTLT1A`, Veggies = `_VEGLT1A`,
         Exercise = `_TOTINDA`, Depression = ADDEPEV3) %>% 
  mutate(Heart_Disease = recode(Heart_Disease, '2' = 0), 
         Depression = recode(Depression, '2' = 0, '7' = as.numeric(NA), '9' = as.numeric(NA)),
         Ment_Health = recode(Ment_Health, '88' = 0, '77' = as.numeric(NA), '99' = as.numeric(NA)),
         Gen_Health = recode(Gen_Health, '2' = 0, '3' = as.numeric(NA), '9' = as.numeric(NA)),
         Phys_Health = recode(Phys_Health, '88' = 0, '77' = as.numeric(NA), '99' = as.numeric(NA)),
         Health_Cov = recode(Health_Cov, '2' = 0, '9' = as.numeric(NA)),
         Med_Cost = recode(Med_Cost, '2' = 0,
                           '7' = as.numeric(NA), '9' = as.numeric(NA)),
         BP = recode(BP, '1' = 0, '2' = 1, '9' = as.numeric(NA)),
         Chol = recode(Chol, '1' = 0, '2' = 1, '9' = as.numeric(NA)),
         Stroke = recode(Stroke, '1' = 1, '2' = 0,
                         '7' = as.numeric(NA), '9' = as.numeric(NA)),
         Diabetes = recode(Diabetes, '2' = 2, '3' = 0, '4' = 1,
                           '1' = 2, '7' = as.numeric(NA),
                           '9' = as.numeric(NA)),
         Education = recode(Education, '9' = as.numeric(NA)),
         Income = recode(Income, '77' = as.numeric(NA), '99' = as.numeric(NA)),
         BMI = BMI/100,
         Smoking = recode(Smoking, '4' = 1, '3' = 2, '2' = 3, '1' = 4, '9' = as.numeric(NA)),
         Avg_Drinks = recode(Avg_Drinks,'88' = 0, '77' = as.numeric(NA), '99' = as.numeric(NA)),
         Exercise = recode(Exercise, '2' = 0, '9' = as.numeric(NA)),
         Sodium_Intake = recode(Sodium_Intake, '2' = 0, '7' = as.numeric(NA), '9' = as.numeric(NA)),
         Fruit = recode(Fruit, '2' = 0, '9' = as.numeric(NA)),
         Veggies = recode(Veggies, '2' = 0, '9' = as.numeric(NA)))

glimpse(survey1)
```

## Variable Descriptions

Heart_Disease: Respondents that have ever reported having coronary heart disease (CHD) or myocardial infarction (MI) \
0 - did not report \
1 - reported \
NA - not asked or missing

Depression: (Ever told) you that you have a depressive disorder, including depression, major depression, dysthymia, or minor depression? \
0 - no \
1 - yes \
NA - don't know, refused \

Ment_Health: Now thinking about your mental health, which includes stress, depression, and problems with emotions, for how many days during the past 30 days was your mental health not good? \
0:30 - # of days \
NA - don't know, refused

Gen_Health: Adults with good or better general health \
0 - fair or poor health \
1 - good or better health \
NA - don't know, not sure or refused/missing

Phys_Health: Now thinking about your physical health, which includes physical illness and injury, for how many days during the past 30 days was your physical health not good? \
0:30 - # days \
NA - don't know, refused, missing

Health_Cov: Respondents aged 18-64 who have any form of health care coverage \
0 - do not have \
1 - have \
NA - don't know, refused, missing

Med_Cost: Was there a time in the past 12 months when you needed to see a doctor but could not because of cost? \
0 - no \
1 - yes \
NA - don't know, refused, missing

BP: Adults who have been told they have high blood pressure by a doctor, nurse, or other health professional \
0 - no \
1 - yes \
NA - don't know, refused, missing

Chol: Adults who have had their cholesterol checked and have been told by a doctor, nurse, or other health professional that it was high \
0 - no \
1 - yes \
NA - don't know, refused, missing

Stroke: (Ever told) you had a stroke \
0 - no \
1 - yes \
NA - don't know, refused

Diabetes: (Ever told) you have diabetes (If "Yes" and respondent is female, ask "Was this only when you were pregnant?". If Respondent says pre-diabetes or borderline diabetes, use response code 4.) \
0 - no \
1 - pre-diabetes or borderline \
2 - yes \
NA - don't know, refused, missing

Sex: Indicate sex of respondent \
1 - male \
2 - female

Education: Level of education completed \
1 - did not graduate highschool \
2 - graduated highschool \
3 - attended college or technical school \
4 - graduated from college or technical school \
NA - don't know, refused, missing

Income: Is your annual household income from all sources: (If respondent refuses at any income level, code "Refused.") \
1 - <$10,000 \
2 - $10,000:$14,999 \
3 - $15,000:$19,999 \
4 - $20,000:$24,999 \
5 - $25,000:$34,999 \
6 - $35,000:$49,999 \
7 - $50,000:$74,999 \
8 - $75,000:$99,999 \
9 - $100,000:$149,999 \
10 - $150,000:$199,000 \
11 - $200,000+ \
NA - don't know, refused, missing

BMI: Body Mass Index (BMI) \
1:99.99 - bmi\
NA - don't know, refused, missing \

Smoking: Four-level smoker status: Everyday smoker, Someday smoker, Former smoker, Non-smoker \
1 - never smoked \
2 - former smoker \
3 - now smokes some days \
4 - now smokes every day \
NA - don't know, refused, missing

Avg_Drinks: One drink is equivalent to a 12-ounce beer, a 5-ounce glass of wine, or a drink with one shot of liquor. During the past 30 days, on the days when you drank, about how many drinks did you drink on the average? (A 40 ounce beer would count as 3 drinks, or a cocktail drink with 2 shots would count as 2 drinks.) \
1:76 - # of drinks \
NA - don't know, refused, missing

Exercise: Adults who reported doing physical activity or exercise during the past 30 days other than their regular job \
0 - did not have physical activity or exercise \
1 - had physical activity or exercise \
NA - don't know, refused, missing

Sodium_Intake: Has a doctor or other health professional ever advised you to reduce sodium or salt intake? \
0 - no \
1 - yes \
NA - don't know, refused, missing

Age_65: Age categories \
1 - age 18:24 \
2 - age 25:34 \
3 - age 35:44 \
4 - age 45:54 \
5 - age 55:64 \
6 - age 65+ \
NA - don't know, refused, missing

Fruit: Consume Fruit 1 or more times per day \
0 - less than one time per day \
1 - one or more times per day \
NA - don't know, refused, missing

Veggies: Consume Vegetables 1 or more times per day \
0 - less than one time per day \
1 - one or more times per day \
NA - don't know, refused, missing

## Exploratory Data Analysis
```{r, warning=FALSE}
# Missingness
colSums(is.na(survey1)) / nrow(survey1)

# Drop variables with >80% missingness
survey2 <- survey1 %>% dplyr::select(-Sodium_Intake)

# Impute variables with sequential, random hot-deck imputation
survey2 <- hotdeck(survey2)
survey2 <- survey2 %>% dplyr::select(1:21)

# Class imbalance
sum(survey2$Heart_Disease) / nrow(survey2)

# Checking for duplicates in original dataset
dat$SEQNO <- as.factor(dat$SEQNO)
duplicates <- dat %>% group_by(SEQNO, `_STATE`) %>%
  summarize(count = n()) %>% filter(count > 1)
dat2 <- dat %>% dplyr::select(-c(`_STATE`:`_PSU`))
table(sapply(dat2, class))
# No duplicates found
```

## Unsupervised Analysis
```{r}
# Hierarchical Clustering

# Use 10000 random rows for clustering
set.seed(100)
survey_cluster <- survey2[sample(nrow(survey2), 10000),]
dis <- dist(survey_cluster[,-1])

# Hierarchical clustering with complete linkage
set.seed(100)
hc <- hclust(dis, method = 'complete')

# Plotting
plot(hc)
rect.hclust(hc, 4)

# Divide into 4 clusters
hc_2 <- cutree(hc, 4)

# Create dataframe with clusters and diagnosis for each observation
disease <- survey_cluster %>% pull(Heart_Disease)
comp <- data.frame(hc_2, disease)

# Create table of amount of each diagnosis for each cluster
with(comp, table(hc_2, disease))

# K-Means Clustering

set.seed(100)
km <- kmeans(survey_cluster[,-1], 2, nstart = 100)
disease <- survey_cluster %>% pull(Heart_Disease)
comp2 <- data.frame(km$cluster, disease)
with(comp2, table(km$cluster, disease))
```

## Relationship Between Mental Health Variables and Heart Disease
```{r}
# Chi-square test of independence between Depression and Heart_Disease variables
# Performed on non-imputed dataset
table(survey1$Heart_Disease, survey1$Depression)
chisq.test(survey1$Heart_Disease, survey1$Depression)
# Chi-square statistic has an associated p-value below 0.05
# Reject null, evidence supporting the two variables not being independent

# Logistic regression to explore relationship between Depression and Heart_Disease
# Performed on non-imputed dataset
logit1 <- glm(Heart_Disease ~ as.factor(Depression), data = survey1, family = "binomial")
summary(logit1)
# Individuals that have been told they have a depressive disorder have a 0.34 higher log odds of ever having reported coronary heart disease or myocardial infarction compared to individuals that have never been told they have a depressive disorder
# This coefficient has a statistically significant z-statistic with a p-value below 0.05

exp(coef(logit1))
# Individuals that have been told they have a depressive disorder have a 2.8% higher odds of ever having reported coronary heart disease or myocardial infarction compared to individuals that have never been told they have a depressive disorder

# Logistic regression to explore relationship between Ment_Health and Heart_Disease
# Performed on non-imputed dataset
logit2 <- glm(Heart_Disease ~ Ment_Health, data = survey1, family = "binomial")
summary(logit2)
# For every one day increase in mental health being considered not good within the past 30 days, the log odds of an individual ever having reported coronary heart disease or myocardial infarction increases by 0.001
# This coefficient has a statistically significant t-statistic with a p-value below 0.05
exp(coef(logit2))
# For every one day increase in mental health being considered not good within the past 30 days, the odds of an individual ever having reported coronary heart disease or myocardial infarction increases by 0.1%
```

## Logistic Regression for Prediction with Mental Health Variables
```{r}
# Change categorical variables to factor
cols <- c("Depression", "Gen_Health", "Health_Cov", "Med_Cost", "BP", "Chol", "Stroke",
          "Diabetes", "Sex", "Education", "Income", "Smoking", "Exercise",
          "Age_Groups", "Fruit", "Veggies")
survey2[cols] = lapply(survey2[cols], factor)

# Split dataset into training and testing data
set.seed(100)
survey3 <- survey2 %>% mutate(Split = sample.split(pull(survey2, 1), SplitRatio = 0.8))
train.set <- survey3 %>% filter(Split == TRUE) %>% dplyr::select(-Split)
test.set.x <- survey3 %>% filter(Split == FALSE) %>% 
  dplyr::select(-c(Split, Heart_Disease))
test.set.y <- survey3 %>% filter(Split == FALSE) %>% dplyr::select(-Split) %>% 
  dplyr::select(Heart_Disease)

# Fit logistic regression model to training data
glm.fit = glm(Heart_Disease ~ ., data=train.set, family=binomial)
summary(glm.fit)
coef(glm.fit)
exp(coef(glm.fit))

# Predict outcome in test set
predicted = glm.fit %>% predict(test.set.x, type = "response")
predicted.classes = ifelse(predicted > 0.5, 1, 0)

# Creating prediction object
pred <- prediction(predicted, test.set.y)

# ROC curve using performance
perf <- performance(pred, "tpr", "fpr")
plot(perf, main = "ROC Curve of Logistic Model with All Variables", 
     sub = "AUC = 0.84")

# Calculating and printing area under the ROC curve
perf_auc <- performance(pred, measure = "auc")
print(perf_auc@y.values)

# Creating cross table
test.set.y$Heart_Disease <- factor(test.set.y$Heart_Disease, c(levels = 0:1),
                                   labels = c("No Heart Disease", "Heart Disease"))
predicted.classes <- factor(predicted.classes, c(levels = 0:1),
                            labels = c("No Heart Disease", "Heart Disease"))
cross <- CrossTable(test.set.y$Heart_Disease, predicted.classes, prop.chisq = FALSE,
                    dnn = c("Observed", "Predicted"))

# Calculating Precision
precision <- cross$prop.col[4]
precision

# Calculating Recall
recall <- cross$prop.row[4]
recall

# Calculating F1 Score
F1 <- (2 * (precision * recall)) / (precision + recall)
F1

# With CV
# glm.fit.cv <- cv.glm(train.set, glm.fit, K = 10)

# Adjusted cross validation estimate of prediction error
# glm.fit.cv$delta[2]
```

## Logistic Regression for Prediction without Mental Health Variables
```{r}
# Split dataset into training and testing data
set.seed(100)
survey3 <- survey2 %>% dplyr::select(-c(Depression, Ment_Health)) %>% 
  mutate(Split = sample.split(pull(survey2, 1), SplitRatio = 0.8))
train.set <- survey3 %>% filter(Split == TRUE) %>% dplyr::select(-Split)
test.set.x <- survey3 %>% filter(Split == FALSE) %>% 
  dplyr::select(-c(Split, Heart_Disease))
test.set.y <- survey3 %>% filter(Split == FALSE) %>% dplyr::select(-Split) %>% 
  dplyr::select(Heart_Disease)

# Fit logistic regression model to training data
glm.fit2 = glm(Heart_Disease ~ ., data=train.set, family=binomial)
summary(glm.fit2)
coef(glm.fit2)
exp(coef(glm.fit2))

# Predict outcome in test set
predicted = glm.fit2 %>% predict(test.set.x, type = "response")
predicted.classes = ifelse(predicted > 0.5, 1, 0)

# Creating prediction object
pred <- prediction(predicted, test.set.y)

# ROC curve using performance
perf <- performance(pred, "tpr", "fpr")
plot(perf, main = "ROC Curve of Logistic Model without Mental Health Variables",
           sub = "AUC = 0.84")

# Calculating and printing area under the ROC curve
perf_auc <- performance(pred, measure = "auc")
print(perf_auc@y.values)

# Creating cross table
test.set.y$Heart_Disease <- factor(test.set.y$Heart_Disease, c(levels = 0:1),
                                   labels = c("No Heart Disease", "Heart Disease"))
predicted.classes <- factor(predicted.classes, c(levels = 0:1),
                            labels = c("No Heart Disease", "Heart Disease"))
cross <- CrossTable(test.set.y$Heart_Disease, predicted.classes, prop.chisq = FALSE,
                    dnn = c("Observed", "Predicted"))

# Calculating Precision
precision <- cross$prop.col[4]
precision

# Calculating Recall
recall <- cross$prop.row[4]
recall

# Calculating F1 Score
F1 <- (2 * (precision * recall)) / (precision + recall)
F1

# With CV
# glm.fit.cv <- cv.glm(train.set, glm.fit, K = 10)

# Adjusted cross validation estimate of prediction error
# glm.fit.cv$delta[2]
```

## Stepwise Logistic
```{r}
# Performing stepwise selection on logistic model using AIC
set.seed(100)
step.model <- glm.fit %>% stepAIC(trace = FALSE)
coef(step.model)
exp(coef(step.model))

# Predict outcome in test set
predicted = step.model %>% predict(test.set.x, type = "response")
predicted.classes = ifelse(predicted > 0.5, 1, 0)

# Creating prediction object
test.set.y <- test.set.y %>% 
  mutate(Heart_Disease = ifelse(Heart_Disease == "No Heart Disease", 0, 1))
pred <- prediction(predicted, test.set.y)

# ROC curve using performance
perf <- performance(pred, "tpr", "fpr")
roc_plot <- plot(perf, main = "ROC Curve of Stepwise Logistic Model",
     sub = "AUC = 0.84")
roc_plot

# Calculating and printing area under the ROC curve
perf_auc <- performance(pred, measure = "auc")
print(perf_auc@y.values)

# Creating cross table
test.set.y$Heart_Disease <- factor(test.set.y$Heart_Disease, c(levels = 0:1),
                                   labels = c("No Heart Disease", "Heart Disease"))
predicted.classes <- factor(predicted.classes, c(levels = 0:1),
                            labels = c("No Heart Disease", "Heart Disease"))
cross <- CrossTable(test.set.y$Heart_Disease, predicted.classes, prop.chisq = FALSE,
                    dnn = c("Observed", "Predicted"))

# Calculating Precision
precision <- cross$prop.col[4]
precision

# Calculating Recall
recall <- cross$prop.row[4]
recall

# Calculating F1 Score
F1 <- (2 * (precision * recall)) / (precision + recall)
F1

# With CV
# glm.fit.cv <- cv.glm(train.set, glm.fit, K = 10)

# Adjusted cross validation estimate of prediction error
# glm.fit.cv$delta[2]
```